---
title: "Dementia Data Analysis"
author: "Daniel Gallardo-Gómez."
date: '2024-03-11'
output: html_document
---

```{r, results='hide', message=FALSE, warning=FALSE, error=FALSE}
# required libraries
library(tidyverse) # run the two following packages if any problem is found here
library(dplyr) # data wrangling
library(ggplot2) # data plotting and visualisation
library(msm) # multi-state modeling 
library(msm.stacked) # transition probabilities plots
library(elect) # life expectancy estimation
library(ggthemes) # data customisation
library(ggsci) # data customisation
```


### Importing data

Clean dataset from NHATS allowing only forward transitions.

```{r}
data <- readRDS("NHATS_data.Rda")
```


## Data modeling: multi-state survival model

Model including all explored covariates: `age`, `sppb10`, `sex`, `exercise`, `income2`, and `smk_before`. You could rerun the model, or simply import the model from the corresponding repository to save time (<https://github.com/dgalgom/Dementia>), since it is a computationally intense process.

```{r, warning=FALSE, eval=FALSE}
# Specifying a model
## A model is ruled by a transition intensity matrix *Q*
## In our example, there are four possible states through which older adults can move
## Dementia-free, MCI, dementia, and death
## Firstly, we assume that the patient can only advance from states while alive, and die from any state
Q <- rbind(c(0, 0.1, 0.1, 0.1),
           c(0, 0, 0.1, 0.1),
           c(0, 0, 0, 0.1),
           c(0, 0, 0, 0))

# Fitting the model into msm
model <- msm(state ~ age, 
                    subject = id, 
                    data = data,
                    covariates = ~ age + sppb.sens +
                      sex + exercise + income2 + smk_before,
                    qmatrix = Q, 
                    deathexact = 4, 
                    control = list(fnscale = 10000,
                                   maxit = 1000,
                                   trace = 1,
                                   REPORT = 1,
                                   reltol = 1e-8),
                    center = FALSE,
                    death = TRUE)
```

```{r}
# importing model from repository
model <- readRDS("model_msm.Rda")
```


Summary of our model: observe hazard ratios for the selected covariates.

```{r, collapse=TRUE}
model
```


Stacked transition probabilities for older adults aged the mean age of individuals in the NHATS (mean age = `r round(mean(data$age, na.rm = TRUE) + 80.5522656302653, 2)`, SD = `r round(sd(data$age, na.rm = TRUE), 2)`) over 10 years having good and poor physical function.

```{r, collapse=TRUE}
# older men with poor physical function plot 
stacked.plot.msm(model = model, tstart = 0, tforward = 10,
                 covariates = list(age = 0,
                                   sppb10 = 0,
                                   sex = 1,
                                   income2 = 0.5,
                                   exercise = 0.5,
                                   smk_before = 0.5),
                 exclude = "State 4") +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_viridis_d(option = "plasma") +
  labs(fill = "To:",
       title = "Older men with poor physical function") 

# older men with good physical function plot
stacked.plot.msm(model = model, tstart = 0, tforward = 10,
                          covariates = list(age = 0,
                                            sppb10 = 1,
                                            sex = 1,
                                            income2 = 0.5,
                                            exercise = 0.5,
                                            smk_before = 0.5),
                          exclude = "State 4") +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_viridis_d(option = "plasma") +
  labs(fill = "To:",
       title = "Older men with good physical function") 

# older women with poor physical function plot
stacked.plot.msm(model = model, tstart = 0, tforward = 10,
                          covariates = list(age = 0,
                                            sppb10 = 0,
                                            sex = 2,
                                            income2 = 0.5,
                                            exercise = 0.5,
                                            smk_before = 0.5),
                          exclude = "State 4") +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_viridis_d(option = "plasma") +
  labs(fill = "To:",
       title = "Older women with poor physical function") 

# older women with good physical function plot
stacked.plot.msm(model = model, tstart = 0, tforward = 10,
                          covariates = list(age = 0,
                                            sppb10 = 1,
                                            sex = 2,
                                            income2 = 0.5,
                                            exercise = 0.5,
                                            smk_before = 0.5),
                          exclude = "State 4") +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_viridis_d(option = "plasma") +
  labs(fill = "To:",
       title = "Older women with good physical function") 
```


## Life expectancy estimates

Life expectancy in a specified state is defined as the expected remaining number of years in that state and is conditional on current state and age.

For these estimates across lifespan, a loop was created to gather all life expectancy information corresponding to **total life expectancy**, **marginal life expectancy estimates** (from the total LE, how much time would you spend in the different states without taking into account the current state), and **state-specific life expectancy estimates**.

```{r}
# first, create an empty list where all the information will be contained
LE_summary <- list()

# loop to calculate LE estimates for each age point for a specific combination of covariates
for (i in seq(-15.55227, 24.44773, 1)) { # from 65 to 105 years old
  LE_summary[[i+16.55227]] <- elect(model, 
                      b.covariates = list(age = i,
                                          sppb10 = 0, # 0: poor physical function; 1: good physical function
                                          sex = 1, # 1: men; 2: women
                                          income2 = 0.5, # the rest of covariates were adjusted to mean values
                                          exercise = 0.5,
                                          smk_before = 0.5),
                      statedistdata = sddata, 
                      time.scale.msm = "years",
                      h = 0.5, 
                      age.max = age.max, 
                      S = 100, # number of simulations to compute uncertainty 
                      setseed = 2012024)
}
```


We advice save this information corresponding with a specific combination of covariates in an object that could be used next. For example:

```{r, eval=FALSE, collapse=TRUE}
# Imagine that we estimated the LE for older men with poor physical function
mod_sum_males_poor <- LE_summary
# mod_sum_males_good <- LE_summary
# mod_sum_females_poor <- LE_summary
# mod_sum_females_good <- LE_summary

# which information can we find in this object?
# for example, if we want to know the LE estimates for olde men aged 65 and with poor physical function
summary.elect(mod_sum_males_poor[[1]])

# e11 is the remaining number of years in dementia-free state for older adults free of any cognitive condition
# e12 is the remaining number of years living with MCI for older adults living free of any cognitive condition
# e13 is the remaining number of years living with dementia for older adults living free of any cognitive condition
# e22 is the remaining number of years living with MCI for older adults with MCI
# e23 is the remaining number of years living with dementia for older adults with MCI
# e33 is the remaining number of years living with dementia for older adults with dementia
# e1 is the expected number of years free of dementia based on the total LE
# e2 is the expected number of years living with MCI based on the total LE
# e3 is the expected number of years with dementia based on the total LE
# e is the total life expectancy 
```


Extracting data for each of the states (dementia-free, MCI, and dementia). In this case, we extract marginal LEs. 

```{r}
# Data for dementia-free state
free.data <- data.frame(sex = rep(c("Women", "Men"), each = 200, times = 41),
           pp = rep(c("Good Physical Function", "Poor Physical Function"), each = 100, times = 82),
           year = rep(seq(65, 105), each = 400),
           type = "Dementia-free",
           value = c(mod_sum_females_good[[1]]$sim[, 10],
                         mod_sum_females_poor[[1]]$sim[, 10],
                         mod_sum_males_good[[1]]$sim[, 10],
                         mod_sum_males_poor[[1]]$sim[, 10],
                     mod_sum_females_good[[2]]$sim[, 10],
                     mod_sum_females_poor[[2]]$sim[, 10],
                     mod_sum_males_good[[2]]$sim[, 10],
                     mod_sum_males_poor[[2]]$sim[, 10],
                     mod_sum_females_good[[3]]$sim[, 10],
                     mod_sum_females_poor[[3]]$sim[, 10],
                     mod_sum_males_good[[3]]$sim[, 10],
                     mod_sum_males_poor[[3]]$sim[, 10],
                     mod_sum_females_good[[4]]$sim[, 10],
                     mod_sum_females_poor[[4]]$sim[, 10],
                     mod_sum_males_good[[4]]$sim[, 10],
                     mod_sum_males_poor[[4]]$sim[, 10],
                     mod_sum_females_good[[5]]$sim[, 10],
                     mod_sum_females_poor[[5]]$sim[, 10],
                     mod_sum_males_good[[5]]$sim[, 10],
                     mod_sum_males_poor[[5]]$sim[, 10],
                     mod_sum_females_good[[6]]$sim[, 10],
                     mod_sum_females_poor[[6]]$sim[, 10],
                     mod_sum_males_good[[6]]$sim[, 10],
                     mod_sum_males_poor[[6]]$sim[, 10],
                     mod_sum_females_good[[7]]$sim[, 10],
                     mod_sum_females_poor[[7]]$sim[, 10],
                     mod_sum_males_good[[7]]$sim[, 10],
                     mod_sum_males_poor[[7]]$sim[, 10],
                     mod_sum_females_good[[8]]$sim[, 10],
                     mod_sum_females_poor[[8]]$sim[, 10],
                     mod_sum_males_good[[8]]$sim[, 10],
                     mod_sum_males_poor[[8]]$sim[, 10],
                     mod_sum_females_good[[9]]$sim[, 10],
                     mod_sum_females_poor[[9]]$sim[, 10],
                     mod_sum_males_good[[9]]$sim[, 10],
                     mod_sum_males_poor[[9]]$sim[, 10],
                     mod_sum_females_good[[10]]$sim[, 10],
                     mod_sum_females_poor[[10]]$sim[, 10],
                     mod_sum_males_good[[10]]$sim[, 10],
                     mod_sum_males_poor[[10]]$sim[, 10],
                     mod_sum_females_good[[11]]$sim[, 10],
                     mod_sum_females_poor[[11]]$sim[, 10],
                     mod_sum_males_good[[11]]$sim[, 10],
                     mod_sum_males_poor[[11]]$sim[, 10],
                     mod_sum_females_good[[12]]$sim[, 10],
                     mod_sum_females_poor[[12]]$sim[, 10],
                     mod_sum_males_good[[12]]$sim[, 10],
                     mod_sum_males_poor[[12]]$sim[, 10],
                     mod_sum_females_good[[13]]$sim[, 10],
                     mod_sum_females_poor[[13]]$sim[, 10],
                     mod_sum_males_good[[13]]$sim[, 10],
                     mod_sum_males_poor[[13]]$sim[, 10],
                     mod_sum_females_good[[14]]$sim[, 10],
                     mod_sum_females_poor[[14]]$sim[, 10],
                     mod_sum_males_good[[14]]$sim[, 10],
                     mod_sum_males_poor[[14]]$sim[, 10],
                     mod_sum_females_good[[15]]$sim[, 10],
                     mod_sum_females_poor[[15]]$sim[, 10],
                     mod_sum_males_good[[15]]$sim[, 10],
                     mod_sum_males_poor[[15]]$sim[, 10],
                     mod_sum_females_good[[16]]$sim[, 10],
                     mod_sum_females_poor[[16]]$sim[, 10],
                     mod_sum_males_good[[16]]$sim[, 10],
                     mod_sum_males_poor[[16]]$sim[, 10],
                     mod_sum_females_good[[17]]$sim[, 10],
                     mod_sum_females_poor[[17]]$sim[, 10],
                     mod_sum_males_good[[17]]$sim[, 10],
                     mod_sum_males_poor[[17]]$sim[, 10],
                     mod_sum_females_good[[18]]$sim[, 10],
                     mod_sum_females_poor[[18]]$sim[, 10],
                     mod_sum_males_good[[18]]$sim[, 10],
                     mod_sum_males_poor[[18]]$sim[, 10],
                     mod_sum_females_good[[19]]$sim[, 10],
                     mod_sum_females_poor[[19]]$sim[, 10],
                     mod_sum_males_good[[19]]$sim[, 10],
                     mod_sum_males_poor[[19]]$sim[, 10],
                     mod_sum_females_good[[20]]$sim[, 10],
                     mod_sum_females_poor[[20]]$sim[, 10],
                     mod_sum_males_good[[20]]$sim[, 10],
                     mod_sum_males_poor[[20]]$sim[, 10],
                     mod_sum_females_good[[21]]$sim[, 10],
                     mod_sum_females_poor[[21]]$sim[, 10],
                     mod_sum_males_good[[21]]$sim[, 10],
                     mod_sum_males_poor[[21]]$sim[, 10],
                     mod_sum_females_good[[22]]$sim[, 10],
                     mod_sum_females_poor[[22]]$sim[, 10],
                     mod_sum_males_good[[22]]$sim[, 10],
                     mod_sum_males_poor[[22]]$sim[, 10],
                     mod_sum_females_good[[23]]$sim[, 10],
                     mod_sum_females_poor[[23]]$sim[, 10],
                     mod_sum_males_good[[23]]$sim[, 10],
                     mod_sum_males_poor[[23]]$sim[, 10],
                     mod_sum_females_good[[24]]$sim[, 10],
                     mod_sum_females_poor[[24]]$sim[, 10],
                     mod_sum_males_good[[24]]$sim[, 10],
                     mod_sum_males_poor[[24]]$sim[, 10],
                     mod_sum_females_good[[25]]$sim[, 10],
                     mod_sum_females_poor[[25]]$sim[, 10],
                     mod_sum_males_good[[25]]$sim[, 10],
                     mod_sum_males_poor[[25]]$sim[, 10],
                     mod_sum_females_good[[26]]$sim[, 10],
                     mod_sum_females_poor[[26]]$sim[, 10],
                     mod_sum_males_good[[26]]$sim[, 10],
                     mod_sum_males_poor[[26]]$sim[, 10],
                     mod_sum_females_good[[27]]$sim[, 10],
                     mod_sum_females_poor[[27]]$sim[, 10],
                     mod_sum_males_good[[27]]$sim[, 10],
                     mod_sum_males_poor[[27]]$sim[, 10],
                     mod_sum_females_good[[28]]$sim[, 10],
                     mod_sum_females_poor[[28]]$sim[, 10],
                     mod_sum_males_good[[28]]$sim[, 10],
                     mod_sum_males_poor[[28]]$sim[, 10],
                     mod_sum_females_good[[29]]$sim[, 10],
                     mod_sum_females_poor[[29]]$sim[, 10],
                     mod_sum_males_good[[29]]$sim[, 10],
                     mod_sum_males_poor[[29]]$sim[, 10],
                     mod_sum_females_good[[30]]$sim[, 10],
                     mod_sum_females_poor[[30]]$sim[, 10],
                     mod_sum_males_good[[30]]$sim[, 10],
                     mod_sum_males_poor[[30]]$sim[, 10],
                     mod_sum_females_good[[31]]$sim[, 10],
                     mod_sum_females_poor[[31]]$sim[, 10],
                     mod_sum_males_good[[31]]$sim[, 10],
                     mod_sum_males_poor[[31]]$sim[, 10],
                     mod_sum_females_good[[32]]$sim[, 10],
                     mod_sum_females_poor[[32]]$sim[, 10],
                     mod_sum_males_good[[32]]$sim[, 10],
                     mod_sum_males_poor[[32]]$sim[, 10],
                     mod_sum_females_good[[33]]$sim[, 10],
                     mod_sum_females_poor[[33]]$sim[, 10],
                     mod_sum_males_good[[33]]$sim[, 10],
                     mod_sum_males_poor[[33]]$sim[, 10],
                     mod_sum_females_good[[34]]$sim[, 10],
                     mod_sum_females_poor[[34]]$sim[, 10],
                     mod_sum_males_good[[34]]$sim[, 10],
                     mod_sum_males_poor[[34]]$sim[, 10],
                     mod_sum_females_good[[35]]$sim[, 10],
                     mod_sum_females_poor[[35]]$sim[, 10],
                     mod_sum_males_good[[35]]$sim[, 10],
                     mod_sum_males_poor[[35]]$sim[, 10],
                     mod_sum_females_good[[36]]$sim[, 10],
                     mod_sum_females_poor[[36]]$sim[, 10],
                     mod_sum_males_good[[36]]$sim[, 10],
                     mod_sum_males_poor[[36]]$sim[, 10],
                     mod_sum_females_good[[37]]$sim[, 10],
                     mod_sum_females_poor[[37]]$sim[, 10],
                     mod_sum_males_good[[37]]$sim[, 10],
                     mod_sum_males_poor[[37]]$sim[, 10],
                     mod_sum_females_good[[38]]$sim[, 10],
                     mod_sum_females_poor[[38]]$sim[, 10],
                     mod_sum_males_good[[38]]$sim[, 10],
                     mod_sum_males_poor[[38]]$sim[, 10],
                     mod_sum_females_good[[39]]$sim[, 10],
                     mod_sum_females_poor[[39]]$sim[, 10],
                     mod_sum_males_good[[39]]$sim[, 10],
                     mod_sum_males_poor[[39]]$sim[, 10],
                     mod_sum_females_good[[40]]$sim[, 10],
                     mod_sum_females_poor[[40]]$sim[, 10],
                     mod_sum_males_good[[40]]$sim[, 10],
                     mod_sum_males_poor[[40]]$sim[, 10],
                     mod_sum_females_good[[41]]$sim[, 10],
                     mod_sum_females_poor[[41]]$sim[, 10],
                     mod_sum_males_good[[41]]$sim[, 10],
                     mod_sum_males_poor[[41]]$sim[, 10]))


# Data for MCI state
mci.data <- data.frame(sex = rep(c("Women", "Men"), each = 200, times = 41),
                        pp = rep(c("Good Physical Function", "Poor Physical Function"), each = 100, times = 82),
                        year = rep(seq(65, 105), each = 400),
                        type = "MCI",
                        value = c(mod_sum_females_good[[1]]$sim[, 11],
                                  mod_sum_females_poor[[1]]$sim[, 11],
                                  mod_sum_males_good[[1]]$sim[, 11],
                                  mod_sum_males_poor[[1]]$sim[, 11],
                                  mod_sum_females_good[[2]]$sim[, 11],
                                  mod_sum_females_poor[[2]]$sim[, 11],
                                  mod_sum_males_good[[2]]$sim[, 11],
                                  mod_sum_males_poor[[2]]$sim[, 11],
                                  mod_sum_females_good[[3]]$sim[, 11],
                                  mod_sum_females_poor[[3]]$sim[, 11],
                                  mod_sum_males_good[[3]]$sim[, 11],
                                  mod_sum_males_poor[[3]]$sim[, 11],
                                  mod_sum_females_good[[4]]$sim[, 11],
                                  mod_sum_females_poor[[4]]$sim[, 11],
                                  mod_sum_males_good[[4]]$sim[, 11],
                                  mod_sum_males_poor[[4]]$sim[, 11],
                                  mod_sum_females_good[[5]]$sim[, 11],
                                  mod_sum_females_poor[[5]]$sim[, 11],
                                  mod_sum_males_good[[5]]$sim[, 11],
                                  mod_sum_males_poor[[5]]$sim[, 11],
                                  mod_sum_females_good[[6]]$sim[, 11],
                                  mod_sum_females_poor[[6]]$sim[, 11],
                                  mod_sum_males_good[[6]]$sim[, 11],
                                  mod_sum_males_poor[[6]]$sim[, 11],
                                  mod_sum_females_good[[7]]$sim[, 11],
                                  mod_sum_females_poor[[7]]$sim[, 11],
                                  mod_sum_males_good[[7]]$sim[, 11],
                                  mod_sum_males_poor[[7]]$sim[, 11],
                                  mod_sum_females_good[[8]]$sim[, 11],
                                  mod_sum_females_poor[[8]]$sim[, 11],
                                  mod_sum_males_good[[8]]$sim[, 11],
                                  mod_sum_males_poor[[8]]$sim[, 11],
                                  mod_sum_females_good[[9]]$sim[, 11],
                                  mod_sum_females_poor[[9]]$sim[, 11],
                                  mod_sum_males_good[[9]]$sim[, 11],
                                  mod_sum_males_poor[[9]]$sim[, 11],
                                  mod_sum_females_good[[10]]$sim[, 11],
                                  mod_sum_females_poor[[10]]$sim[, 11],
                                  mod_sum_males_good[[10]]$sim[, 11],
                                  mod_sum_males_poor[[10]]$sim[, 11],
                                  mod_sum_females_good[[11]]$sim[, 11],
                                  mod_sum_females_poor[[11]]$sim[, 11],
                                  mod_sum_males_good[[11]]$sim[, 11],
                                  mod_sum_males_poor[[11]]$sim[, 11],
                                  mod_sum_females_good[[12]]$sim[, 11],
                                  mod_sum_females_poor[[12]]$sim[, 11],
                                  mod_sum_males_good[[12]]$sim[, 11],
                                  mod_sum_males_poor[[12]]$sim[, 11],
                                  mod_sum_females_good[[13]]$sim[, 11],
                                  mod_sum_females_poor[[13]]$sim[, 11],
                                  mod_sum_males_good[[13]]$sim[, 11],
                                  mod_sum_males_poor[[13]]$sim[, 11],
                                  mod_sum_females_good[[14]]$sim[, 11],
                                  mod_sum_females_poor[[14]]$sim[, 11],
                                  mod_sum_males_good[[14]]$sim[, 11],
                                  mod_sum_males_poor[[14]]$sim[, 11],
                                  mod_sum_females_good[[15]]$sim[, 11],
                                  mod_sum_females_poor[[15]]$sim[, 11],
                                  mod_sum_males_good[[15]]$sim[, 11],
                                  mod_sum_males_poor[[15]]$sim[, 11],
                                  mod_sum_females_good[[16]]$sim[, 11],
                                  mod_sum_females_poor[[16]]$sim[, 11],
                                  mod_sum_males_good[[16]]$sim[, 11],
                                  mod_sum_males_poor[[16]]$sim[, 11],
                                  mod_sum_females_good[[17]]$sim[, 11],
                                  mod_sum_females_poor[[17]]$sim[, 11],
                                  mod_sum_males_good[[17]]$sim[, 11],
                                  mod_sum_males_poor[[17]]$sim[, 11],
                                  mod_sum_females_good[[18]]$sim[, 11],
                                  mod_sum_females_poor[[18]]$sim[, 11],
                                  mod_sum_males_good[[18]]$sim[, 11],
                                  mod_sum_males_poor[[18]]$sim[, 11],
                                  mod_sum_females_good[[19]]$sim[, 11],
                                  mod_sum_females_poor[[19]]$sim[, 11],
                                  mod_sum_males_good[[19]]$sim[, 11],
                                  mod_sum_males_poor[[19]]$sim[, 11],
                                  mod_sum_females_good[[20]]$sim[, 11],
                                  mod_sum_females_poor[[20]]$sim[, 11],
                                  mod_sum_males_good[[20]]$sim[, 11],
                                  mod_sum_males_poor[[20]]$sim[, 11],
                                  mod_sum_females_good[[21]]$sim[, 11],
                                  mod_sum_females_poor[[21]]$sim[, 11],
                                  mod_sum_males_good[[21]]$sim[, 11],
                                  mod_sum_males_poor[[21]]$sim[, 11],
                                  mod_sum_females_good[[22]]$sim[, 11],
                                  mod_sum_females_poor[[22]]$sim[, 11],
                                  mod_sum_males_good[[22]]$sim[, 11],
                                  mod_sum_males_poor[[22]]$sim[, 11],
                                  mod_sum_females_good[[23]]$sim[, 11],
                                  mod_sum_females_poor[[23]]$sim[, 11],
                                  mod_sum_males_good[[23]]$sim[, 11],
                                  mod_sum_males_poor[[23]]$sim[, 11],
                                  mod_sum_females_good[[24]]$sim[, 11],
                                  mod_sum_females_poor[[24]]$sim[, 11],
                                  mod_sum_males_good[[24]]$sim[, 11],
                                  mod_sum_males_poor[[24]]$sim[, 11],
                                  mod_sum_females_good[[25]]$sim[, 11],
                                  mod_sum_females_poor[[25]]$sim[, 11],
                                  mod_sum_males_good[[25]]$sim[, 11],
                                  mod_sum_males_poor[[25]]$sim[, 11],
                                  mod_sum_females_good[[26]]$sim[, 11],
                                  mod_sum_females_poor[[26]]$sim[, 11],
                                  mod_sum_males_good[[26]]$sim[, 11],
                                  mod_sum_males_poor[[26]]$sim[, 11],
                                  mod_sum_females_good[[27]]$sim[, 11],
                                  mod_sum_females_poor[[27]]$sim[, 11],
                                  mod_sum_males_good[[27]]$sim[, 11],
                                  mod_sum_males_poor[[27]]$sim[, 11],
                                  mod_sum_females_good[[28]]$sim[, 11],
                                  mod_sum_females_poor[[28]]$sim[, 11],
                                  mod_sum_males_good[[28]]$sim[, 11],
                                  mod_sum_males_poor[[28]]$sim[, 11],
                                  mod_sum_females_good[[29]]$sim[, 11],
                                  mod_sum_females_poor[[29]]$sim[, 11],
                                  mod_sum_males_good[[29]]$sim[, 11],
                                  mod_sum_males_poor[[29]]$sim[, 11],
                                  mod_sum_females_good[[30]]$sim[, 11],
                                  mod_sum_females_poor[[30]]$sim[, 11],
                                  mod_sum_males_good[[30]]$sim[, 11],
                                  mod_sum_males_poor[[30]]$sim[, 11],
                                  mod_sum_females_good[[31]]$sim[, 11],
                                  mod_sum_females_poor[[31]]$sim[, 11],
                                  mod_sum_males_good[[31]]$sim[, 11],
                                  mod_sum_males_poor[[31]]$sim[, 11],
                                  mod_sum_females_good[[32]]$sim[, 11],
                                  mod_sum_females_poor[[32]]$sim[, 11],
                                  mod_sum_males_good[[32]]$sim[, 11],
                                  mod_sum_males_poor[[32]]$sim[, 11],
                                  mod_sum_females_good[[33]]$sim[, 11],
                                  mod_sum_females_poor[[33]]$sim[, 11],
                                  mod_sum_males_good[[33]]$sim[, 11],
                                  mod_sum_males_poor[[33]]$sim[, 11],
                                  mod_sum_females_good[[34]]$sim[, 11],
                                  mod_sum_females_poor[[34]]$sim[, 11],
                                  mod_sum_males_good[[34]]$sim[, 11],
                                  mod_sum_males_poor[[34]]$sim[, 11],
                                  mod_sum_females_good[[35]]$sim[, 11],
                                  mod_sum_females_poor[[35]]$sim[, 11],
                                  mod_sum_males_good[[35]]$sim[, 11],
                                  mod_sum_males_poor[[35]]$sim[, 11],
                                  mod_sum_females_good[[36]]$sim[, 11],
                                  mod_sum_females_poor[[36]]$sim[, 11],
                                  mod_sum_males_good[[36]]$sim[, 11],
                                  mod_sum_males_poor[[36]]$sim[, 11],
                                  mod_sum_females_good[[37]]$sim[, 11],
                                  mod_sum_females_poor[[37]]$sim[, 11],
                                  mod_sum_males_good[[37]]$sim[, 11],
                                  mod_sum_males_poor[[37]]$sim[, 11],
                                  mod_sum_females_good[[38]]$sim[, 11],
                                  mod_sum_females_poor[[38]]$sim[, 11],
                                  mod_sum_males_good[[38]]$sim[, 11],
                                  mod_sum_males_poor[[38]]$sim[, 11],
                                  mod_sum_females_good[[39]]$sim[, 11],
                                  mod_sum_females_poor[[39]]$sim[, 11],
                                  mod_sum_males_good[[39]]$sim[, 11],
                                  mod_sum_males_poor[[39]]$sim[, 11],
                                  mod_sum_females_good[[40]]$sim[, 11],
                                  mod_sum_females_poor[[40]]$sim[, 11],
                                  mod_sum_males_good[[40]]$sim[, 11],
                                  mod_sum_males_poor[[40]]$sim[, 11],
                                  mod_sum_females_good[[41]]$sim[, 11],
                                  mod_sum_females_poor[[41]]$sim[, 11],
                                  mod_sum_males_good[[41]]$sim[, 11],
                                  mod_sum_males_poor[[41]]$sim[, 11]))

# Data for dementia state
dem.data <- data.frame(sex = rep(c("Women", "Men"), each = 200, times = 41),
                       pp = rep(c("Good Physical Function", "Poor Physical Function"), each = 100, times = 82),
                       year = rep(seq(65, 105), each = 400),
                       type = "Dementia",
                       value = c(mod_sum_females_good[[1]]$sim[, 12],
                                 mod_sum_females_poor[[1]]$sim[, 12],
                                 mod_sum_males_good[[1]]$sim[, 12],
                                 mod_sum_males_poor[[1]]$sim[, 12],
                                 mod_sum_females_good[[2]]$sim[, 12],
                                 mod_sum_females_poor[[2]]$sim[, 12],
                                 mod_sum_males_good[[2]]$sim[, 12],
                                 mod_sum_males_poor[[2]]$sim[, 12],
                                 mod_sum_females_good[[3]]$sim[, 12],
                                 mod_sum_females_poor[[3]]$sim[, 12],
                                 mod_sum_males_good[[3]]$sim[, 12],
                                 mod_sum_males_poor[[3]]$sim[, 12],
                                 mod_sum_females_good[[4]]$sim[, 12],
                                 mod_sum_females_poor[[4]]$sim[, 12],
                                 mod_sum_males_good[[4]]$sim[, 12],
                                 mod_sum_males_poor[[4]]$sim[, 12],
                                 mod_sum_females_good[[5]]$sim[, 12],
                                 mod_sum_females_poor[[5]]$sim[, 12],
                                 mod_sum_males_good[[5]]$sim[, 12],
                                 mod_sum_males_poor[[5]]$sim[, 12],
                                 mod_sum_females_good[[6]]$sim[, 12],
                                 mod_sum_females_poor[[6]]$sim[, 12],
                                 mod_sum_males_good[[6]]$sim[, 12],
                                 mod_sum_males_poor[[6]]$sim[, 12],
                                 mod_sum_females_good[[7]]$sim[, 12],
                                 mod_sum_females_poor[[7]]$sim[, 12],
                                 mod_sum_males_good[[7]]$sim[, 12],
                                 mod_sum_males_poor[[7]]$sim[, 12],
                                 mod_sum_females_good[[8]]$sim[, 12],
                                 mod_sum_females_poor[[8]]$sim[, 12],
                                 mod_sum_males_good[[8]]$sim[, 12],
                                 mod_sum_males_poor[[8]]$sim[, 12],
                                 mod_sum_females_good[[9]]$sim[, 12],
                                 mod_sum_females_poor[[9]]$sim[, 12],
                                 mod_sum_males_good[[9]]$sim[, 12],
                                 mod_sum_males_poor[[9]]$sim[, 12],
                                 mod_sum_females_good[[10]]$sim[, 12],
                                 mod_sum_females_poor[[10]]$sim[, 12],
                                 mod_sum_males_good[[10]]$sim[, 12],
                                 mod_sum_males_poor[[10]]$sim[, 12],
                                 mod_sum_females_good[[11]]$sim[, 12],
                                 mod_sum_females_poor[[11]]$sim[, 12],
                                 mod_sum_males_good[[11]]$sim[, 12],
                                 mod_sum_males_poor[[11]]$sim[, 12],
                                 mod_sum_females_good[[12]]$sim[, 12],
                                 mod_sum_females_poor[[12]]$sim[, 12],
                                 mod_sum_males_good[[12]]$sim[, 12],
                                 mod_sum_males_poor[[12]]$sim[, 12],
                                 mod_sum_females_good[[13]]$sim[, 12],
                                 mod_sum_females_poor[[13]]$sim[, 12],
                                 mod_sum_males_good[[13]]$sim[, 12],
                                 mod_sum_males_poor[[13]]$sim[, 12],
                                 mod_sum_females_good[[14]]$sim[, 12],
                                 mod_sum_females_poor[[14]]$sim[, 12],
                                 mod_sum_males_good[[14]]$sim[, 12],
                                 mod_sum_males_poor[[14]]$sim[, 12],
                                 mod_sum_females_good[[15]]$sim[, 12],
                                 mod_sum_females_poor[[15]]$sim[, 12],
                                 mod_sum_males_good[[15]]$sim[, 12],
                                 mod_sum_males_poor[[15]]$sim[, 12],
                                 mod_sum_females_good[[16]]$sim[, 12],
                                 mod_sum_females_poor[[16]]$sim[, 12],
                                 mod_sum_males_good[[16]]$sim[, 12],
                                 mod_sum_males_poor[[16]]$sim[, 12],
                                 mod_sum_females_good[[17]]$sim[, 12],
                                 mod_sum_females_poor[[17]]$sim[, 12],
                                 mod_sum_males_good[[17]]$sim[, 12],
                                 mod_sum_males_poor[[17]]$sim[, 12],
                                 mod_sum_females_good[[18]]$sim[, 12],
                                 mod_sum_females_poor[[18]]$sim[, 12],
                                 mod_sum_males_good[[18]]$sim[, 12],
                                 mod_sum_males_poor[[18]]$sim[, 12],
                                 mod_sum_females_good[[19]]$sim[, 12],
                                 mod_sum_females_poor[[19]]$sim[, 12],
                                 mod_sum_males_good[[19]]$sim[, 12],
                                 mod_sum_males_poor[[19]]$sim[, 12],
                                 mod_sum_females_good[[20]]$sim[, 12],
                                 mod_sum_females_poor[[20]]$sim[, 12],
                                 mod_sum_males_good[[20]]$sim[, 12],
                                 mod_sum_males_poor[[20]]$sim[, 12],
                                 mod_sum_females_good[[21]]$sim[, 12],
                                 mod_sum_females_poor[[21]]$sim[, 12],
                                 mod_sum_males_good[[21]]$sim[, 12],
                                 mod_sum_males_poor[[21]]$sim[, 12],
                                 mod_sum_females_good[[22]]$sim[, 12],
                                 mod_sum_females_poor[[22]]$sim[, 12],
                                 mod_sum_males_good[[22]]$sim[, 12],
                                 mod_sum_males_poor[[22]]$sim[, 12],
                                 mod_sum_females_good[[23]]$sim[, 12],
                                 mod_sum_females_poor[[23]]$sim[, 12],
                                 mod_sum_males_good[[23]]$sim[, 12],
                                 mod_sum_males_poor[[23]]$sim[, 12],
                                 mod_sum_females_good[[24]]$sim[, 12],
                                 mod_sum_females_poor[[24]]$sim[, 12],
                                 mod_sum_males_good[[24]]$sim[, 12],
                                 mod_sum_males_poor[[24]]$sim[, 12],
                                 mod_sum_females_good[[25]]$sim[, 12],
                                 mod_sum_females_poor[[25]]$sim[, 12],
                                 mod_sum_males_good[[25]]$sim[, 12],
                                 mod_sum_males_poor[[25]]$sim[, 12],
                                 mod_sum_females_good[[26]]$sim[, 12],
                                 mod_sum_females_poor[[26]]$sim[, 12],
                                 mod_sum_males_good[[26]]$sim[, 12],
                                 mod_sum_males_poor[[26]]$sim[, 12],
                                 mod_sum_females_good[[27]]$sim[, 12],
                                 mod_sum_females_poor[[27]]$sim[, 12],
                                 mod_sum_males_good[[27]]$sim[, 12],
                                 mod_sum_males_poor[[27]]$sim[, 12],
                                 mod_sum_females_good[[28]]$sim[, 12],
                                 mod_sum_females_poor[[28]]$sim[, 12],
                                 mod_sum_males_good[[28]]$sim[, 12],
                                 mod_sum_males_poor[[28]]$sim[, 12],
                                 mod_sum_females_good[[29]]$sim[, 12],
                                 mod_sum_females_poor[[29]]$sim[, 12],
                                 mod_sum_males_good[[29]]$sim[, 12],
                                 mod_sum_males_poor[[29]]$sim[, 12],
                                 mod_sum_females_good[[30]]$sim[, 12],
                                 mod_sum_females_poor[[30]]$sim[, 12],
                                 mod_sum_males_good[[30]]$sim[, 12],
                                 mod_sum_males_poor[[30]]$sim[, 12],
                                 mod_sum_females_good[[31]]$sim[, 12],
                                 mod_sum_females_poor[[31]]$sim[, 12],
                                 mod_sum_males_good[[31]]$sim[, 12],
                                 mod_sum_males_poor[[31]]$sim[, 12],
                                 mod_sum_females_good[[32]]$sim[, 12],
                                 mod_sum_females_poor[[32]]$sim[, 12],
                                 mod_sum_males_good[[32]]$sim[, 12],
                                 mod_sum_males_poor[[32]]$sim[, 12],
                                 mod_sum_females_good[[33]]$sim[, 12],
                                 mod_sum_females_poor[[33]]$sim[, 12],
                                 mod_sum_males_good[[33]]$sim[, 12],
                                 mod_sum_males_poor[[33]]$sim[, 12],
                                 mod_sum_females_good[[34]]$sim[, 12],
                                 mod_sum_females_poor[[34]]$sim[, 12],
                                 mod_sum_males_good[[34]]$sim[, 12],
                                 mod_sum_males_poor[[34]]$sim[, 12],
                                 mod_sum_females_good[[35]]$sim[, 12],
                                 mod_sum_females_poor[[35]]$sim[, 12],
                                 mod_sum_males_good[[35]]$sim[, 12],
                                 mod_sum_males_poor[[35]]$sim[, 12],
                                 mod_sum_females_good[[36]]$sim[, 12],
                                 mod_sum_females_poor[[36]]$sim[, 12],
                                 mod_sum_males_good[[36]]$sim[, 12],
                                 mod_sum_males_poor[[36]]$sim[, 12],
                                 mod_sum_females_good[[37]]$sim[, 12],
                                 mod_sum_females_poor[[37]]$sim[, 12],
                                 mod_sum_males_good[[37]]$sim[, 12],
                                 mod_sum_males_poor[[37]]$sim[, 12],
                                 mod_sum_females_good[[38]]$sim[, 12],
                                 mod_sum_females_poor[[38]]$sim[, 12],
                                 mod_sum_males_good[[38]]$sim[, 12],
                                 mod_sum_males_poor[[38]]$sim[, 12],
                                 mod_sum_females_good[[39]]$sim[, 12],
                                 mod_sum_females_poor[[39]]$sim[, 12],
                                 mod_sum_males_good[[39]]$sim[, 12],
                                 mod_sum_males_poor[[39]]$sim[, 12],
                                 mod_sum_females_good[[40]]$sim[, 12],
                                 mod_sum_females_poor[[40]]$sim[, 12],
                                 mod_sum_males_good[[40]]$sim[, 12],
                                 mod_sum_males_poor[[40]]$sim[, 12],
                                 mod_sum_females_good[[41]]$sim[, 12],
                                 mod_sum_females_poor[[41]]$sim[, 12],
                                 mod_sum_males_good[[41]]$sim[, 12],
                                 mod_sum_males_poor[[41]]$sim[, 12]))


# Combining datasets
comp.data <- free.data %>%
  rbind(mci.data, dem.data)


# plot!
comp.data %>%
  group_by(sex, pp, year, type) %>%
  summarise(value = mean(value)) %>%
  ungroup() %>%
  tidyr::spread(type, value) %>% # transform long to wide format data
  select(sex, pp, year, `Dementia-free`, MCI, Dementia) %>%
  mutate(prop.free = (`Dementia-free` / (`Dementia-free` + MCI + Dementia)),
         prop.mci = (MCI / (`Dementia-free` + MCI + Dementia)),
         prop.dem = 1 - prop.free - prop.mci) %>%
  select(sex, pp, year, prop.free, prop.mci, prop.dem) %>%
  rename(`Dementia-free` = prop.free,
         MCI = prop.mci,
         Dementia = prop.dem) %>%
  tidyr::gather(key = "type", value = "value", -c(sex, pp, year)) # reshaping data
%>% 
  mutate(type = factor(type,
                       levels = c("Dementia",
                                  "MCI",
                                  "Dementia-free"))) %>%
  ggplot(aes(fill = type, x = year, y = value)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(pp ~ sex) +
  labs(x = "Age (years)",
       y = "Proportion of Total Life Expectancy",
       fill = "") +
  theme_base() +
  scale_x_continuous(breaks = seq(65, 105, 10)) +
  scale_y_continuous(limits = seq(0, 100),
                     labels = scales::percent_format()) +
  theme(legend.position = "bottom")+
  scale_fill_brewer(breaks = c("Dementia-free",
                               "MCI",
                               "Dementia"),
                    palette = "Greys")
```


Also, we outputted the numerical data of this plot.

```{r}
# Numerical data corresponding to total LE and marginal LEs
comp.data %>%
  group_by(sex, pp, year, type) %>%
  summarise(value = mean(value)) %>%
  ungroup() %>%
  tidyr::spread(type, value) %>%
  select(sex, pp, year, `Dementia-free`, MCI, Dementia) %>%
  mutate(prop.free = (`Dementia-free` / (`Dementia-free` + MCI + Dementia)) * 100,
         prop.dem = (Dementia / (`Dementia-free` + MCI + Dementia)) * 100,
         prop.mci = 100 - prop.free - prop.dem) %>% 
  mutate_if(is.numeric, round, 2) %>%
  mutate(pp = ifelse(pp == "Good Physical Function", "Good", "Poor")) %>%
  select(sex, pp, year, `Dementia-free`, MCI, Dementia, prop.free, prop.mci, prop.dem)
```


State-specific life expectancy estimates. 

```{r}
# Define the index for the LE estimation you want to extract
i <- 1 # for life expectancy free of dementia in older adults free of any cognitive condition
# i <- 2 # for example, it would correspond with the life expectancy with MCI in older adults free of any cognitive condition

# model 1 for comparison; for example, men with good physical function
x <- mod_sum_males_good

# model 2 for comparison; for example, men with poor physical function to compare good vs. poor PF in men
y <- mod_sum_males_poor

# calculation of mean differences
males_good_males_poor <- as.data.frame(mean_qi(x[[1]]$sim[, i] - y[[1]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[2]]$sim[, i] - y[[2]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[3]]$sim[, i] - y[[3]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[4]]$sim[, i] - y[[4]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[5]]$sim[, i] - y[[5]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[6]]$sim[, i] - y[[6]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[7]]$sim[, i] - y[[7]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[8]]$sim[, i] - y[[8]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[9]]$sim[, i] - y[[9]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[10]]$sim[, i] - y[[10]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[11]]$sim[, i] - y[[11]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[12]]$sim[, i] - y[[12]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[13]]$sim[, i] - y[[13]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[14]]$sim[, i] - y[[14]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[15]]$sim[, i] - y[[15]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[16]]$sim[, i] - y[[16]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[17]]$sim[, i] - y[[17]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[18]]$sim[, i] - y[[18]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[19]]$sim[, i] - y[[19]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[20]]$sim[, i] - y[[20]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[21]]$sim[, i] - y[[21]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[22]]$sim[, i] - y[[22]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[23]]$sim[, i] - y[[23]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[24]]$sim[, i] - y[[24]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[25]]$sim[, i] - y[[25]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[26]]$sim[, i] - y[[26]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[27]]$sim[, i] - y[[27]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[28]]$sim[, i] - y[[28]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[29]]$sim[, i] - y[[29]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[30]]$sim[, i] - y[[30]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[31]]$sim[, i] - y[[31]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[32]]$sim[, i] - y[[32]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[33]]$sim[, i] - y[[33]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[34]]$sim[, i] - y[[34]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[35]]$sim[, i] - y[[35]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[36]]$sim[, i] - y[[36]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[37]]$sim[, i] - y[[37]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[38]]$sim[, i] - y[[38]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[39]]$sim[, i] - y[[39]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[40]]$sim[, i] - y[[40]]$sim[, i])[, 1:3]) %>%
  rbind(mean_qi(x[[41]]$sim[, i] - y[[41]]$sim[, i])[, 1:3]) %>%
  mutate_if(is.numeric, round, 2) %>%
  cbind(year = seq(65, 105),
        comparison = rep("Men with good PF\nvs.\nMen with poor PF"))

# plot for this comparison
males_good_males_poor %>%
  # rbind(females_good_females_poor) %>% # just in case you want to combine men and women plots
  # mutate(comparison = ifelse(comparison == "Women with good PF\nvs.\nWomen with poor PF", "Women", "Men")) %>%
  ggplot(aes(x = year, y = y)) +
  geom_pointinterval(aes(ymin = ymin,
                         ymax = ymax)) +
  # facet_wrap(~ comparison) +
  labs(y = "Estimated years of life free of dementia gained with a good physical function",
       x = "Age (years)") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(65, 105, 10)) +
  scale_y_continuous(breaks = seq(-7, 20, 0.5)) +
  theme(axis.title.y = element_text(size = 9,
                                    vjust = +3),
        axis.title.x = element_text(vjust = -0.75))
```


Additionally, we computed the numerical value of these differences.

```{r}
males_good_males_poor %>%
# rbind(females_good_females_poor, # just in case of data combination
#        males_good_males_poor) %>% 
   mutate(comparison = ifelse(comparison == "Women with good PF\nvs.\nWomen with poor PF", "Women", "Men"),
         `Current state` = "Dementia",
         `Expected number of years in this state` = "Dementia") %>%
  dplyr::select(`Current state`, `Expected number of years in this state`,
                comparison, year, y, ymin, ymax)
```


